Comparison of linear beam theories

### 2024-03-16 article.tex
fixe

\section{Cantilever two-dimensional model}\label{ssec:2D_Model:FEM}
Let $\{e_1,e_2\}$ be a right-handed orthonormal basis for $R^2$. Denote the elastic body by $\Omega \in R^2$, with $(0,0)$ as the point of reference. For a rectangular cross-section, the body $\Omega$ can be described as,
\begin{eqnarray*}
	\Omega = \left \{ x \in E_2 \ | \ 0 \leq x_1 \leq 1, \ -\frac{h}{2} \leq x_2 \leq \frac{h}{2} \right \},
\end{eqnarray*} where $\partial \Omega$ denotes the boundary of $\Omega$. The boundary $\partial \Omega$ can be divided into the four distinct lines (or edges) as follows: \label{sym:En}

\noindent\begin{minipage}{.5\linewidth}
	\begin{eqnarray*}
		\Sigma:& \quad x_1 &= 0\\
		\Gamma_3:& \quad x_1 &= 1
	\end{eqnarray*}
\end{minipage}%
\begin{minipage}{.5\linewidth}
	\begin{eqnarray*}
		\Gamma_1:& \quad x_2 &= -{h}/{2}\\
		\Gamma_2:& \quad x_2 &= {h}/{2}
	\end{eqnarray*}
\end{minipage}

In this configuration, the body is clamped rigidly to the surface at $x_1 = 0$ denoted as $\Sigma$ and the body is free-hanging on the other boundaries denoted by $\Gamma$. This is a catilever configuration.

@@

The variational problem for the two dimensional cantilever model is also derived in \ref{ssec:2D_Model} and refered to as Problem 2DV. Using the reference configuration, the variational problem can be rewritten into a model specific form.

\subsubsection{Problem 2DV}\label{sssec:2D_Model:Problem2D1VChp5}
Find a function $u$ such that for all $t>0$, $u \in T(\Omega)$ and 
\begin{align}
	c(u,\phi) = -b(u,\phi) + (Q,\phi) \label{eq:2D_Model:Problem2D1VEqChp5}
\end{align}
for all $\phi \in T(\Omega)$. With the test function space 
\begin{eqnarray*}
	T(\Omega) & = & \left\{ \phi \in C^1(\bar{\Omega}) \ | \ \phi = 0 \ \textrm{ on } \ \Gamma \right\}.
\end{eqnarray*}

@@

\subsubsection{Constitutive equations:}
\begin{eqnarray}
	\sigma_{11} & = & \frac{1}{\gamma(1-\nu^2)}(\partial_1 u_1 + \nu \partial_2 u_2) \label{CE1} \\
	\sigma_{22} & = & \frac{1}{\gamma(1-\nu^2)}(\partial_2 u_2 + \nu \partial_1 u_1) \label{CE2} \\
	\sigma_{12} & = & \frac{1}{2\gamma(1+\nu)}(\partial_1 u_2 + \partial_2 u_1) \label{CE3}
\end{eqnarray}

\subsubsection{Bilinear form:}
\begin{align}
	b(u,\phi) & = & \frac{1}{\gamma(1-\nu^2)}\int_{\Omega} (\partial_1 u_1 \partial_1 \phi_1 + \partial_2 u_2 \partial _2 \phi_2 + \nu\partial_1 u_1 \partial_2\phi_2 + \nu \partial_2 u_2 \partial_1 \phi_1 ) \ dA \nonumber\\
	& + & \frac{1}{2\gamma(1+\nu)}\int_{\Omega} (\partial_1 u_2 \partial_1 \phi_2 + \partial_1 u_2 \partial_2 \phi_1 + \partial_2 u_1 \partial_1\phi_2 + \partial_2 u_1 \partial_2\phi_1) \ dA.
\end{align}

\subsection{Galerkin approximation}\label{2d_FEM_G}
To be able to apply the Finite Element Method to Problem 2DV, the body $\Omega$ needs to be discretised. This is done by dividing the body $\Omega$ into discrete shapes, called elements. There are various types of shapes of elements that can be used. Since the body $\omega$ is has a rectangular cross-section, rectangular elements are the simplest elements to use.

Divide the reference configuration $\Omega$ into a grid of rectangular elements, such that there are $n = n_1\times n_2$ nodes.

Define a set of $n$-dimensional linear independent basis functions. For the two dimensional model, the basis functions can be defined by the set $$B = \left\{\langle\phi_1, 0\rangle, \langle\phi_2, 0\rangle,...,\langle\phi_{n}, 0 \rangle,\langle 0,\phi_1\rangle,\langle 0 ,\phi_2\rangle,...,\langle 0,\phi_{n}\rangle \right\}.$$

These functions are chosen as piecewise Hermite bi-cubic functions $\phi_i$. Simpler bi-linear functions can also be used, however the use of the bi-cubic basis functions results in faster convergence and also the benefit of obtaining the derivatives of the solution at the expense of more complexity.

The subset of basis functions $B$ that satisfies all the conditions of the test function space $T(\Omega)$ are called the admissible basis functions. Denote the admissible basis functions by $\delta_j$, where each $\delta_j$ is a unique element of $B$. The admissible basis functions can be numbered and expressed as the set $A = \left \{\delta_1, \delta_2,..., \delta_k \right\}$ for some $k \leq 2n$.\\

@@

\subsubsection{Problem 2DG}
Find a function $u^h$ such that for all $t>0$, $u^h \in S^h$ and
\begin{eqnarray*}
	(u^h, \phi_i) & = & -b(u^h,\phi_i) + (Q^I, \phi_i)
\end{eqnarray*} for $i = 1,2,...,k$. $Q^I$ is scalar vector obtained after interpolating the function $Q$ over the rectangular grid $\Omega$. i.e. $Q^I_{i,j} = Q(x_i,x_j)$ for $i = 1,2,...,n_1$ and $j = 1,2,...,n_2$.

\subsection{System of differential equations}\label{ssec:2DFEM:DE}

Consider the following standard Finite Element Method matrices
\subsubsection{FEM matrices}
\noindent\begin{minipage}{.5\linewidth}
	\begin{eqnarray*}
		\mathbf{M}_{j,i} & = & \int_{\Omega} \phi_i \phi_j ~dA \\
		\mathbf{{K}_{11}}_{j,i} & = & \int_{\Omega} \partial_1\phi_i \partial_1\phi_j~dA\\
		\mathbf{{K}_{22}}_{j,i} & = & \int_{\Omega} \partial_2\phi_i \partial_2\phi_j~dA
	\end{eqnarray*}
\end{minipage}%
\begin{minipage}{.5\linewidth}
	\begin{eqnarray*}
		\mathbf{{K}_{12}}_{j,i} & = & \int_{\Omega} \partial_2\phi_i \partial_1\phi_j~dA\\
		\mathbf{{K}_{21}}_{j,i} & = & \int_{\Omega} \partial_1\phi_i \partial_2\phi_j~dA
	\end{eqnarray*}
\end{minipage}
for $i,j = 1,2,...,k$.

@@

\subsubsection{Problem 2DODE}
Find a function $\bar{u} \in S^h$ such that
\begin{eqnarray}
	M\ddot{\bar{u}} & = & K\bar{u} + M_{f}Q^I.
\end{eqnarray} With $\bar{u}$ in the form $\bar{u} = \langle u, \partial_1 u, \partial_2 u, \partial_{12} u \rangle$.

@@

\subsection{Eigenvalue problem}\label{2dFEM_EP}
For the eigenvalue problem, assume that there is no external force, $M_{f}Q^I = 0$, so that 
\begin{eqnarray}
		M\ddot{\bar{u}} & = & K\bar{u}.\label{eq:2DFEM:M2}
\end{eqnarray}
It is known that a system of ordinary differential equations has a general solution of the form $e^{rt}$. Suppose that $\bar{w} = e^{\lambda t} \bar{u}$ is a solution for \eqref{eq:2DFEM:M2}. In this solution, $\lambda$ is an eigenvalue and $\bar{u}$ a corresponding eigenfunction. Substitution into \eqref{eq:2DFEM:M2} results in
\begin{eqnarray*}
	M\lambda e^{\lambda t}\bar{u} & = & Ke^{\lambda t}\bar{u}.
\end{eqnarray*}
Since $e^{\lambda t} > 0$ for all values of $\lambda t$, we can cancel it from both sides of the equation and formulate the eigenvalue problem Problem 2DE.

\subsubsection{Problem 2DE}
Find a real number $\lambda$ and a function $\bar{u} \in S^h$ such that
\begin{eqnarray}
	M\lambda{\bar{u}} & = & K\bar{u}.
\end{eqnarray}

\section{Cantilever three-dimensional model} \label{ssec:3D_Model:FEM}
Let $\{e_1,e_2,e_3\}$ be a right-handed orthonormal basis for $R^3$. Denote the elastic body as $\Omega \in R^3$ with $(0,0,0)$ the point of reference. For a rectangular cross-section, the body $\Omega$ can be described as
\begin{eqnarray*}
	\Omega = \left\{ x \in R^3 \ | \ 0 \leq x_1 \leq 1, \ -\frac{h}{2} \leq x_2 \leq \frac{h}{2} , \ -\frac{b}{2} \leq x_3 \leq \frac{b}{2}\right \}
\end{eqnarray*}
Let $\partial \Omega$ denote the boundary of the body. Divide $\partial \Omega$ into the six distinct flat surfaces as follows:

\noindent\begin{minipage}{.5\linewidth}
	\begin{eqnarray*}
		\Sigma:& \quad x_1 &= 0\\
		\Gamma_5:& \quad x_1 &= 1\\
		\Gamma_1:& \quad x_3 &= -{b}/{2} 
	\end{eqnarray*}
\end{minipage}%
\begin{minipage}{.5\linewidth}
	\begin{eqnarray*}
		\Gamma_2:& \quad x_2 &= -{h}/{2}\\
		\Gamma_3:& \quad x_3 &= {b}/{2}\\
		\Gamma_4:& \quad x_2 &= {h}/{2} 
	\end{eqnarray*}
\end{minipage}

Using this notation, the boundary conditions are similar to the two-dimensional model, where the body is clamped rigidly at $\Sigma$, and free-hanging on the other sides denoted by $\Gamma$.

\subsubsection{Cantilever elastic body}
Consider a three-dimensional elastic body with rectangular cross-section, rigidly clamped to a surface at attached at the side $\Sigma$ and free-hanging at all the other sides. This is Problem 3D in section \ref{ssec:3D_Model}.

@@

In section \ref{ssec:3D_Model}, the variational problem for the three-dimensional cantilever model is defined by Problem 3DV. This general form can now be rewritten as in a model specific form using the reference configuration.

\subsubsection{Problem 3DV}
Find a function $u$ such that for all $t>0$, $u \in T(\Omega)$ and 
\begin{align}
	c(u,\phi) = -b(u,\phi) - (Q,\phi) \label{3DB_20}
\end{align}
for all $\phi \in T(\Omega)$.\\

@@

\subsubsection{Bilinear Form}
\begin{align*}
	b(u,\phi)  &=  \int_\Omega~c_1\textrm{Tr}({\cal E}\Phi)+ c_2\textrm{Tr}({\cal E})
	\textrm{Tr}(\Phi) ~dV \nonumber\\
	 &=  \int_{\Omega} \sigma_{11} \partial_1\phi_1 + \sigma_{12}\partial_1 \phi_2 + \sigma_{13}\partial_1 \phi_3 + \sigma_{21}\partial_2 \phi_1 + \sigma_{22}\partial_2\phi_2 + \sigma_{23}\partial_2 \phi_3 ~dV
\end{align*}

\subsection{Galerkin approximation}
For this is three-dimensional body, three-dimensional elements shapes are required. The elements used in this dissertation are rectangular prismatic elements. These elements are also known as `brick-shaped' elements and is described in \cite{Wu06}. This shape of element is a natural choice for a three-dimensional elastic body with a rectangular cross-section.

Divide the reference configuration $\Omega$ into a grid of rectangular prismatic elements, such that there are $n = n_1 \times n_2 \times n_3$ nodes.

Define a set of $n$-dimensional linear independent basis functions. For the three-dimensional model, the basis functions can be defined by the set
\begin{eqnarray*}
 B = \{\langle\phi_1, 0 , 0\rangle, \langle\phi_2, 0, 0\rangle,...,\langle\phi_{n}, 0, 0 \rangle,\\
	\langle 0,\phi_1 ,0 \rangle,\langle 0 ,\phi_2,0\rangle,...,\langle 0,\phi_{n},0\rangle,\\
	\langle 0,0,\phi_1 \rangle,\langle 0,0,\phi_2\rangle,...,\langle 0,0,\phi_{n}\rangle \}.
\end{eqnarray*}

For the this three-dimensional model, piecewise Hermite tri-cubic basis functions are used. Although this is more complex than using piecwise tri-linear basis functions, the tri-cubic basis functions ensure faster convergence and also the derivatives of the solutions.

Recall that the admissible basis functions, are all the basis functions that satisfies all the conditions of the test function space $T(\Omega)$. Denote the admissible basis functions by $\delta_j$, with each $\delta_j$ a unique element of $B$. The set of admissible basis functions can then be expressed as $A = \left\{\delta_1, \delta_2,... , \delta_k\right\}$ with $k \leq 3n$.

@@

\subsubsection{Problem 3DG}
Find a function $u^h$ such that for all $t>0$, $u^h \in S^h$ and
\begin{eqnarray*}
	(u^h, \phi_i) & = & -b(u^h,\phi_i) + (Q^I \cdot \phi_i)
\end{eqnarray*} for $i = 1,2,...,k$. $Q^I$ is scalar vector obtained after interpolating the function $Q$ over the rectangular grid $\Omega$. i.e. $Q^I_{i,j, h} = Q(x_i,x_j, x_h)$ for $i = 1,2,...,n_1$, $j = 1,2,...,n_2$ and $h = 1,2,...,n_3$.

\subsection{System of ordinary differential equations}\label{3d_fem_g}
Consider the following standard Finite Element Method matrices.

\subsubsection{FEM matrices}
\noindent\begin{minipage}{.5\linewidth}
	\begin{eqnarray*}
		\mathbf{M}_{ij} & = & \int_{\Omega} \phi_j \phi_i \ dV \ \\
		{K_{11}}_{ij} & = & \int_{\Omega} \partial_1\phi_j \partial_1\phi_i \ dV \  \\
		{K_{12}}_{ij} & = & \int_{\Omega} \partial_1\phi_j \partial_2\phi_i \ dV \ \\
		{K_{13}}_{ij} & = & \int_{\Omega} \partial_1\phi_j \partial_3\phi_i \ dV \ 
	\end{eqnarray*}
\end{minipage}%
\begin{minipage}{.5\linewidth}
	\begin{eqnarray*}
		{K_{22}}_{ij} & = & \int_{\Omega} \partial_2\phi_j \partial_2\phi_i \ dV \  \\
		{K_{23}}_{ij} & = & \int_{\Omega} \partial_2\phi_j \partial_3\phi_i \ dV \ \\
		{K_{33}}_{ij} & = & \int_{\Omega} \partial_3\phi_j \partial_3\phi_i \ dV \ \\
	\end{eqnarray*}
\end{minipage}
for $i,j = 1,2,...,k.$\\

@@

\subsubsection{Problem 3DODE}
Find a function $\bar{u} \in S^h$ such that
\begin{eqnarray}
	M\ddot{\bar{u}} & = & K\bar{u} + M_{f}Q^I. \label{3D_M}
\end{eqnarray} With $\bar{u}$ in the form $\bar{u} = \partial^i_1u \partial^j_2u \partial^k_3u$ for $i,j,k = 0,1,2,3$.

\subsection{Eigenvalue problem}
The derivation and form of the eigenvalue problem for the three-dimensional elastic body is similar to the two-dimensional model given in section \ref{2dFEM_EP}.

\subsubsection{Problem 3DE}\label{3dFEM_EP}
Find a real number $\lambda$ and a function $\bar{u} \in S^h$ such that
\begin{eqnarray}
	M\lambda{\bar{u}} & = & K\bar{u}.
\end{eqnarray}

@@

\section{Validity of a model for a cantilever two-dimensional beam} \label{sec:validity-of-a-2d-beam}
As mentioned in the introduction of the chapter, a beam is a three-dimensional body and therefore a three-dimensional model is more realistic. However the authors of \cite{LVV09} mention that a direct comparison of the one and three-dimensional models will have complexities and they suggest using the two-dimensional model as an intermediate step.

So in this section, the article \cite{LVV09} is extended and the validity of a cantilever two-dimensional beam model is investigated, using a cantilever three-dimensional model as a reference.

\subsection{The models}
Figure \ref{fig:2Dv3D} shows the two beams, Problem 2D and Problem 3D side-by-side.

@@

\subsection{Calculating the eigenvalues}
In section \ref{ssec:2D_Model:FEM} and \ref{ssec:3D_Model:FEM}, the Finite Element Method for the two-dimensional beam and three-dimensional beam is derived. Using this method, the eigenvalue problems can be solved to obtain the eigenvalues and eigenvectors of the two models.

\subsubsection{Problem 2DE and 3DE}
Find a real number $\lambda$ and a function $\bar{u} \in S^h$ such that
\begin{eqnarray}
	K\bar{u} & = & M\lambda{\bar{u}},
\end{eqnarray} where $K$ and $M$ are the standard Finite Element Method matrices defined in section \ref{ssec:2DFEM:DE} for the Problem 2DE and section \ref{3d_fem_g} for Problem 3DE.

\subsubsection{Accuracy of the eigenvalues}
When solving the eigenvalue problems with a numerical method, the accuracy of the eigenvalues can be established by investigating the rate of convergence.

\begin{figure}[H]
    \centering
    \begin{adjustbox}{center}
        \includegraphics[scale=0.7]{files/Chapter4/Convergence.png}
    \end{adjustbox}
    \caption{Rate of convergence of the first 20 eigenvalues for Problem 2DE.}
    \label{fig:conv_2d_eig}
\end{figure}

\begin{figure}[H]
    \centering
    \begin{adjustbox}{center}
        \includegraphics[scale=0.7]{files/Chapter6/convergence3d.png}
    \end{adjustbox}
    \caption{Rate of convergence of the first 20 eigenvalues for Problem 3DE.}
    \label{fig:conv_3d_eig}
\end{figure}

Figure \ref{fig:conv_2d_eig} and \ref{fig:conv_3d_eig} shows the rate of convergence of the first 20 eigenvalues of the two-dimensional beam and the three-dimensional beam respectively. Each color represents a spesific eigenvalue.

These figures show that the two-dimensional model has a much faster rate of convergence than the three-dimensional model. The two-dimensional model also converge faster to a higher accuracy than the three-dimensional model.

Using this, the number of elements can be chosen so that at least the first 10 eigenvalues are accurate to 5 significant digits for the two-dimensional model. For the three-dimensional model, obtaining this level of accuracy can be difficult and computationally expensive, but it is possible to get 4 significant digits of accuracy for the first 10 eigenvalues.

\subsection{Comparing the mode shapes}
To be able to compare the eigenvalues, the mode shapes of the two models are compared to match up the eigenvalues of the two models. This is the same approach as in the article \cite{LVV09}.

When calculating the mode shapes, distinct patterns are observed. There are mode shapes that are present in both models, mode shapes that are only present in the three-dimensional model. The shared mode shapes also include mode shapes that relate to beam problems. These are the mode shapes that are of interest the article \cite{LVV09}.

The focus of the investigation of this article is also on beam type problems. Below are some examples of the mode shapes for beam type eigenvalues, mode shapes for non-beam type eigenvalues that are shared between the two and three-dimensional models and also mode shapes for non-beam type eigenvalues that are only present in the three-dimensional model.

\subsubsection{Mode shapes relating to beam type eigenvalues.}
Figure \ref{fig:beam-2dv3d} show some examples of beam type mode shapes for the displacement $u$.

@@

\subsubsection{Mode shapes relating to non-beam type eigenvalues that are present in the two-dimensional model.}
Figure \ref{fig:nonbeam-2dv3d} show examples of mode shapes relating to non-beam type eigenvalues for the displacement $u$.

@@

\subsection{Comparing the eigenvalues}
For a realistic comparison of the models, the parameters need to be chosen carefully. The parameters are $h$ representing the height of the beam and $b$ representing the width of the beam. The two-dimensional model does not have the width parameter.

For $h$, values are chosen that covers a range of beam shapes from a short thick beam, to a long slender beam. two cases are selected that represents realistic cases. For a short and thick beam, consider $h = 1/5$ and for a long and slender beam, consider $h = 1/20$.

For the parameter $b$, two different cases will be considered. The first case is for $b \leq h$, and the $b > h$. The distinction of these two cases will become apparent in the results. It is important to note that the parameter $b$ will be expressed as a multiple of $h$.

All of the results will include all the eigenvalues shared between the two-models, including not beam type eigenvalues. The non-beam type eigenvalues will be highlighted in grey. The non-beam type eigenvalues that are not shared between the two models will be excluded from the results, but the numbering of the eigenvalues will be kept as if they were included.

>>>>

\section{Cantilever two-dimensional model}\label{ssec:2D_Model:FEM}
Let $\{e_1,e_2\}$ be a right-handed orthonormal basis for $R^2$. Denote the elastic body by $\Omega \in R^2$, with $(0,0)$ as the point of reference. For a rectangular cross-section, the body $\Omega$ can be described as,
\begin{eqnarray*}
	\Omega = \left \{ x \in E_2 \ | \ 0 \leq x_1 \leq 1, \ -\frac{h}{2} \leq x_2 \leq \frac{h}{2} \right \},
\end{eqnarray*} where $\partial \Omega$ denotes the boundary of $\Omega$.

@@

 Using the reference configuration, the variational problem Problem 2DV can be rewritten into a model specific form.

\subsubsection*{Problem 2DV}\label{sssec:2D_Model:Problem2D1VChp5}
Find a function $u$ such that for all $t>0$, $u \in T(\Omega)$ and 
\begin{align}
	c(u,\phi) = -b(u,\phi) + (Q,\phi) \label{eq:2D_Model:Problem2D1VEqChp5}
\end{align}
for all $\phi \in T(\Omega)$. With the test function space 
\begin{eqnarray*}
	T(\Omega) & = & \left\{ \phi \in C^1(\bar{\Omega}) \ | \ \phi = 0 \ \textrm{ on } \ \Gamma \right\}.
\end{eqnarray*}

@@

\subsubsection*{Constitutive equations:}
\begin{eqnarray}
	\sigma_{11} & = & \frac{1}{\gamma(1-\nu^2)}(\partial_1 u_1 + \nu \partial_2 u_2) \label{CE1} \\
	\sigma_{22} & = & \frac{1}{\gamma(1-\nu^2)}(\partial_2 u_2 + \nu \partial_1 u_1) \label{CE2} \\
	\sigma_{12} & = & \frac{1}{2\gamma(1+\nu)}(\partial_1 u_2 + \partial_2 u_1) \label{CE3}
\end{eqnarray}

\subsubsection*{Bilinear form:}
\begin{align}
	b(u,\phi) & = & \frac{1}{\gamma(1-\nu^2)}\int_{\Omega} (\partial_1 u_1 \partial_1 \phi_1 + \partial_2 u_2 \partial _2 \phi_2 + \nu\partial_1 u_1 \partial_2\phi_2 + \nu \partial_2 u_2 \partial_1 \phi_1 ) \ dA \nonumber\\
	& + & \frac{1}{2\gamma(1+\nu)}\int_{\Omega} (\partial_1 u_2 \partial_1 \phi_2 + \partial_1 u_2 \partial_2 \phi_1 + \partial_2 u_1 \partial_1\phi_2 + \partial_2 u_1 \partial_2\phi_1) \ dA.
\end{align}

\subsection*{Galerkin approximation}\label{2d_FEM_G}
To apply the Finite Element Method to Problem 2DV, $\Omega$ must be discretized by dividing it into rectangular elements, suitable for its rectangular cross-section. This process involves creating a grid with $n = n_1 \times n_2$ nodes across $\Omega$.

A set of $n$-dimensional linear independent basis functions is defined, particularly for the two-dimensional model, as 
$$B = \{\langle\phi_1, 0\rangle, \langle\phi_2, 0\rangle,...,\langle\phi_{n}, 0 \rangle,\langle 0,\phi_1\rangle,\langle 0 ,\phi_2\rangle,...,\langle 0,\phi_{n}\rangle \}.$$ 
These functions, chosen as piecewise Hermite bi-cubic functions $\phi_i$, offer quicker convergence and the ability to derive solution derivatives, despite their increased complexity compared to bi-linear functions.

The subset of basis functions from $B$ that meet the criteria of the test function space $T(\Omega)$ are termed admissible basis functions, denoted by $\delta_j$, where each $\delta_j$ is a distinct element of $B$. These can be enumerated and presented as the set 
$$A = \{\delta_1, \delta_2,..., \delta_k \}$$ 
for some $k \leq 2n$.

@@

\subsubsection*{Problem 2DG}
Find a function $u^h$ such that for all $t>0$, $u^h \in S^h$ and
\begin{eqnarray*}
	(u^h, \phi_i) & = & -b(u^h,\phi_i) + (Q^I, \phi_i)
\end{eqnarray*} for $i = 1,2,...,k$. $Q^I$ is scalar vector obtained after interpolating the function $Q$ over the rectangular grid $\Omega$. i.e. $Q^I_{i,j} = Q(x_i,x_j)$ for $i = 1,2,...,n_1$ and $j = 1,2,...,n_2$.

Consider the following standard Finite Element Method matrices

\noindent\begin{minipage}{.5\linewidth}
	\begin{eqnarray*}
		\mathbf{M}_{j,i} & = & \int_{\Omega} \phi_i \phi_j ~dA \\
		\mathbf{{K}_{11}}_{j,i} & = & \int_{\Omega} \partial_1\phi_i \partial_1\phi_j~dA\\
		\mathbf{{K}_{22}}_{j,i} & = & \int_{\Omega} \partial_2\phi_i \partial_2\phi_j~dA
	\end{eqnarray*}
\end{minipage}%
\begin{minipage}{.5\linewidth}
	\begin{eqnarray*}
		\mathbf{{K}_{12}}_{j,i} & = & \int_{\Omega} \partial_2\phi_i \partial_1\phi_j~dA\\
		\mathbf{{K}_{21}}_{j,i} & = & \int_{\Omega} \partial_1\phi_i \partial_2\phi_j~dA
	\end{eqnarray*}
\end{minipage}
for $i,j = 1,2,...,k$.

@@

\subsubsection*{Problem 2DODE}
Find a function $\bar{u} \in S^h$ such that
\begin{eqnarray}
	M\ddot{\bar{u}} & = & K\bar{u} + M_{f}Q^I.
\end{eqnarray} With $\bar{u}$ in the form $\bar{u} = \langle u, \partial_1 u, \partial_2 u, \partial_{12} u \rangle$.

@@

\subsection*{Eigenvalue problem}\label{2dFEM_EP}
For the eigenvalue problem, assume that there is no external force, $M_{f}Q^I = 0$, so that 
\begin{eqnarray}
		M\ddot{\bar{u}} & = & K\bar{u}.\label{eq:2DFEM:M2}
\end{eqnarray}
It is known that a system of ordinary differential equations has a general solution of the form $e^{rt}$. Suppose that $\bar{w} = e^{\lambda t} \bar{u}$ is a solution for \eqref{eq:2DFEM:M2}. In this solution, $\lambda$ is an eigenvalue and $\bar{u}$ a corresponding eigenfunction. Substitution into \eqref{eq:2DFEM:M2} results in
\begin{eqnarray*}
	M\lambda e^{\lambda t}\bar{u} & = & Ke^{\lambda t}\bar{u}.
\end{eqnarray*}
Since $e^{\lambda t} > 0$ for all values of $\lambda t$, we can cancel it from both sides of the equation and formulate the eigenvalue problem Problem 2DE.

\subsubsection*{Problem 2DE}
Find a real number $\lambda$ and a function $\bar{u} \in S^h$ such that
\begin{eqnarray}
	M\lambda{\bar{u}} & = & K\bar{u}.
\end{eqnarray}

\section{Cantilever three-dimensional model} \label{ssec:3D_Model:FEM}
Let $\{e_1,e_2,e_3\}$ be a right-handed orthonormal basis for $R^3$. Denote the elastic body as $\Omega \in R^3$ with $(0,0,0)$ the point of reference. For a rectangular cross-section, the body $\Omega$ can be described as
\begin{eqnarray*}
	\Omega = \left\{ x \in R^3 \ | \ 0 \leq x_1 \leq 1, \ -\frac{h}{2} \leq x_2 \leq \frac{h}{2} , \ -\frac{b}{2} \leq x_3 \leq \frac{b}{2}\right \}
\end{eqnarray*}
Let $\partial \Omega$ denote the boundary of the body.

\FloatBarrier
\begin{figure}[h!]
	\centering
	\begin{tikzpicture}

@@

This general variational form Problem 3DV can now be rewritten as in a model specific form using the reference configuration.

\subsubsection*{Problem 3DV}
Find a function $u$ such that for all $t>0$, $u \in T(\Omega)$ and 
\begin{align}
	c(u,\phi) = -b(u,\phi) - (Q,\phi) \label{3DB_20}
\end{align}
for all $\phi \in T(\Omega)$.\\

@@

\subsubsection*{Bilinear Form}
\begin{align*}
	b(u,\phi)  &=  \int_\Omega~c_1\textrm{Tr}({\cal E}\Phi)+ c_2\textrm{Tr}({\cal E})
	\textrm{Tr}(\Phi) ~dV \nonumber\\
	 &=  \int_{\Omega} \sigma_{11} \partial_1\phi_1 + \sigma_{12}\partial_1 \phi_2 + \sigma_{13}\partial_1 \phi_3 + \sigma_{21}\partial_2 \phi_1 + \sigma_{22}\partial_2\phi_2 + \sigma_{23}\partial_2 \phi_3 ~dV
\end{align*}

\subsection*{Galerkin approximation}
For analyzing a three-dimensional body, rectangular prismatic (brick-shaped) elements are used, as detailed in \cite{Wu06}. This choice suits bodies with rectangular cross-sections.

Split $\Omega$ into a grid of these elements, creating $n = n_1 \times n_2 \times n_3$ nodes.

Define $n$-dimensional linear independent basis functions for the three-dimensional model as
\begin{eqnarray*}
 B = \{\langle\phi_1, 0 , 0\rangle, \langle\phi_2, 0, 0\rangle,...,\langle\phi_{n}, 0, 0 \rangle,\\
	\langle 0,\phi_1 ,0 \rangle,\langle 0 ,\phi_2,0\rangle,...,\langle 0,\phi_{n},0\rangle,\\
	\langle 0,0,\phi_1 \rangle,\langle 0,0,\phi_2\rangle,...,\langle 0,0,\phi_{n}\rangle \}.
\end{eqnarray*}
Employing piecewise Hermite tri-cubic basis functions offers quicker convergence and solution derivatives, albeit with increased complexity compared to tri-linear functions.

Admissible basis functions, satisfying all conditions of $T(\Omega)$, are denoted by $\delta_j$, making up the set $A = \left\{\delta_1, \delta_2,... , \delta_k\right\}$, where $k \leq 3n$.

@@

\subsubsection*{Problem 3DG}
Find a function $u^h$ such that for all $t>0$, $u^h \in S^h$ and
\begin{eqnarray*}
	(u^h, \phi_i) & = & -b(u^h,\phi_i) + (Q^I \cdot \phi_i)
\end{eqnarray*} for $i = 1,2,...,k$. $Q^I$ is scalar vector obtained after interpolating the function $Q$ over the rectangular grid $\Omega$. i.e. $Q^I_{i,j, h} = Q(x_i,x_j, x_h)$ for $i = 1,2,...,n_1$, $j = 1,2,...,n_2$ and $h = 1,2,...,n_3$.

Consider the following standard Finite Element Method matrices.

\noindent\begin{minipage}{.5\linewidth}
	\begin{eqnarray*}
		\mathbf{M}_{ij} & = & \int_{\Omega} \phi_j \phi_i \ dV \ \\
		{K_{11}}_{ij} & = & \int_{\Omega} \partial_1\phi_j \partial_1\phi_i \ dV \  \\
		{K_{12}}_{ij} & = & \int_{\Omega} \partial_1\phi_j \partial_2\phi_i \ dV \ \\
		{K_{13}}_{ij} & = & \int_{\Omega} \partial_1\phi_j \partial_3\phi_i \ dV \ 
	\end{eqnarray*}
\end{minipage}%
\begin{minipage}{.5\linewidth}
	\begin{eqnarray*}
		{K_{22}}_{ij} & = & \int_{\Omega} \partial_2\phi_j \partial_2\phi_i \ dV \  \\
		{K_{23}}_{ij} & = & \int_{\Omega} \partial_2\phi_j \partial_3\phi_i \ dV \ \\
		{K_{33}}_{ij} & = & \int_{\Omega} \partial_3\phi_j \partial_3\phi_i \ dV \ \\
	\end{eqnarray*}
\end{minipage}
for $i,j = 1,2,...,k.$\\

@@

\subsubsection*{Problem 3DODE}
Find a function $\bar{u} \in S^h$ such that
\begin{eqnarray}
	M\ddot{\bar{u}} & = & K\bar{u} + M_{f}Q^I. \label{3D_M}
\end{eqnarray} With $\bar{u}$ in the form $\bar{u} = \partial^i_1u \partial^j_2u \partial^k_3u$ for $i,j,k = 0,1,2,3$.

\subsection*{Eigenvalue problem}
The derivation and form of the eigenvalue problem for the three-dimensional elastic body is similar to the two-dimensional model given in section \ref{2dFEM_EP}.

\subsubsection*{Problem 3DE}\label{3dFEM_EP}
Find a real number $\lambda$ and a function $\bar{u} \in S^h$ such that
\begin{eqnarray}
	M\lambda{\bar{u}} & = & K\bar{u}.
\end{eqnarray}

@@

\section{Validity of a model for a cantilever two-dimensional beam} \label{sec:validity-of-a-2d-beam}
In this section, we compare the two-dimensional cantilever beam with a three-dimensional cantilever beam.

Figure \ref{fig:2Dv3D} shows the two beams, Problem 2D and Problem 3D side-by-side.

@@

\subsection{Calculating the eigenvalues and mode shapes}
The eigenvalues and mode shapes of the two models are calculated using the computer program. The eigenvalues and mode shapes are then compared to determine the accuracy of the two-dimensional model for the cantilever beam. The accuracy of the eigenvalues depends on the number of elements employed in the finite element method. For the two-dimensional model, accuracy reaches at least five significant digits for the first 10 eigenvalues, while the three-dimensional model achieves at least four significant digits for its first 10 eigenvalues.

\subsection{Comparing the mode shapes}
Following \cite{LVV09}, the mode shapes of the two models are compared. As seen in \cite{LVV09}, the eigenvalues of the Timoshenko beam model (beam-type eigenvalues) are all present in the two-dimensional model. However not all the eigenvalues of the two-dimensional model are present in the Timoshenko beam model. These are called non-beam type eigenvalues.

The same is observed for the three-dimensional model, as there are eigenvalues that are present in the three-dimensional model that are not present in the two-dimensional model.

Therefore it is required to match the eigenvalues of the two models by comparing the mode shapes.

\subsubsection*{Mode shapes relating to beam type eigenvalues.}
Figure \ref{fig:beam-2dv3d} show some examples of beam type mode shapes for the displacement $u$.

@@

\subsubsection*{Mode shapes relating to non-beam type eigenvalues that are present in the two-dimensional model.}
Figure \ref{fig:nonbeam-2dv3d} show examples of mode shapes relating to non-beam type eigenvalues for the displacement $u$.

@@

\subsection{Comparing the eigenvalues}
For a realistic comparison of the models, it's crucial to carefully select parameters $h$ and $b$, representing the beam's height and width, respectively, noting that the two-dimensional model lacks the width parameter $b$.

Values for $h$ span a range from short, thick beams ($h = 1/5$) to long, slender beams ($h = 1/20$), capturing two realistic scenarios. Regarding $b$, it's considered in two scenarios: when $b \leq h$ and when $b > h$, with $b$ expressed as a multiple of $h$. This distinction influences the results.

Results will cover all shared eigenvalues between the models, highlighting non-beam type eigenvalues in grey. Non-beam type eigenvalues unique to either model are excluded, but eigenvalue numbering will proceed as if they were included.
